---
title: "Lung Cancer Data Exploratory Analysis"
output: html_notebook
---

# Introduction

Team Member: Jasmine, Vasant, Cheng Lang

This notebook performs an exploratory data analysis (EDA) on lung cancer incidence data in the United States, covering years 2000 to 2019.  
The data comes from multiple CSV files where we combined and cleaned for analysis.  
We will explore the structure of the data set, clean the data, and visualize some key variables to understand the trends and distributions when applicable.

# Package Set Up

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(GGally)
```

# Combining files together (all csv has the exact same data structure, varied by year)
```{r message=FALSE, warning=FALSE}
files <- list.files("data", pattern = "\\.csv$", full.names = TRUE, recursive = TRUE, ignore.case = TRUE) %>% discard(~ grepl("codebook", .x, ignore.case = TRUE))
files
```


```{r message=FALSE, warning=FALSE}
lung_cancer_df <- files %>%
  map_dfr(read_csv) %>%
  clean_names()
```

```{r}
glimpse(lung_cancer_df)
```


```{r message=FALSE, warning=FALSE}
colSums(is.na(lung_cancer_df))
```
```{r}
skim(lung_cancer_df)
```

```{r}
codebook <- read_csv("data/codebook.csv")
glimpse(codebook)
skim(codebook)
```
```{r}
unique(lung_cancer_df$metric_name) # Mortality rate is the only measure here
```






